---
title: "PhD Chapter Summary"
author: "Ross Whippo"
format: html
editor: visual
---

## Chapter 1 - Antarctic Algal Fatty Acids

### Summary

Macroalgae synthesize fatty acids (FA) that higher tropic levels consume to maintain their fundamental biological processes.  The FA content of Antarctic algae have only been described for a limited number of groups including 16 species from 11 families. As part of an NSF-funded project aimed at describing the macroalgal communities of Antarctica, I will be analysing a subset of  algal tissues collected during the project to describe the fatty-acid profiles of 32 species, half of which have no published FA profiles. 

### Methods

Field methods Algal samples were collected from the Western Antarctic Peninsula (WAP) from April-May 2019 via SCUBA at 13 sites spanning approximately 5° of latitude. Algal tissues were collected by hand between 5-40m depth and frozen at -80° C for transport back to the lab for extraction and analysis.

### Fatty acid profiles 

Three representative replicates of each species were extracted from the tissues (sample size allowing) and were processed in a gas chromatograph equipped with a mass spectrometer (GCMS). All chromatograms will be integrated and scored, then processed to identify FA content and analyzed using PERMANOVA with the 'vegan' package in R to detect differences among species. Once all samples are analysed, I will summarise their profiles for publication as a fundamental analysis of nearshore Antarctic trophic ecology. 
 
### Results



## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
